---
title: "Target-Distractor-Chance Comparisons"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

#set seed
```

## Intro

A quick simulation to demonstrate that comparing %target choices to chance is not equivalent to comparing % target choices to %distractor choices (where %distractor = 1 - %target).

We'll quickly simulate 30 participants who complete 10 trials where they can choose either the target or distractor, assuming a true probability of p = 0.6 for choosing the target.

```{r cars}
n_part <- 30
n_trials <- 10
true_probability <- 0.6

#use the binomial distribution to simulate the number of "successes"/ target choices
num_target_choices <- rbinom(n_part,n_trials,true_probability)


d <- data.frame(
  participant_id = seq(1,n_part),
  n_trials = rep(n_trials,n_part),
  target_choices = num_target_choices
) %>%
  mutate(
    distractor_choices = n_trials-target_choices,
    target_prop = target_choices/n_trials,
    distractor_prop = distractor_choices/n_trials
  )
```

## Plot

```{r}
ggplot(d, aes(target_prop))+
  geom_histogram()+
  geom_density()+
  geom_vline(xintercept=0.5)+
  geom_vline(xintercept=0.6,linetype="dashed")
```

## T-Test against chance

Here are the results for the t-test comparing proportion target choices to chance (0.5)

```{r}
t.test(d$target_prop,mu=0.5)
```

## Paired t-test comparing proportion target choices to proportion distractor choices

```{r}
t.test(d$target_prop,d$distractor_prop,paired=TRUE)
```

